---
title: "p8105_hw1_sdc2157"
author: "Stephanie Calluori"
date: 2023-10-04
output: github_document
---

```{r, load_packages, message = FALSE}
library(tidyverse)
library(readxl)
```

# Importing Data and NA Check
```{r, import_data, results = "hide", message = FALSE}
pols_month_raw <- 
  read_csv("data/pols-month.csv")

sum(is.na(pols_month_raw))

snp_raw <-
  read_csv("data/snp.csv")

sum(is.na(snp_raw))

unemployment_raw <-
  read_csv("data/unemployment.csv")

sum(is.na(unemployment_raw))

mr_trash_raw <-
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 1, range = "A2:N586")

sum(is.na(pull(mr_trash_raw, Dumpster)))

prof_trash_raw <-
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 2, range = "A2:M108")

sum(is.na(pull(prof_trash_raw, Dumpster)))

gwyn_trash_raw <-
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = 4, range = "A2:L157")

sum(is.na(pull(gwyn_trash_raw, Dumpster)))

mci_amyloid_raw <-
  read_csv("data/mci_amyloid.csv", skip = 1, col_names = TRUE)

sum(is.na(mci_amyloid_raw))

mci_baseline_raw <- 
  read_csv("data/MCI_baseline.csv", skip = 1, col_names = TRUE)

sum(is.na(mci_baseline_raw))

```

# Problem 1

## Pols-month Cleaning

```{r, pols data exploration, include = FALSE}
pols_month_raw |>
  count(prez_gop)

pols_month_raw |>
  count(prez_dem)
```

```{r, pols cleaning}
month_df = 
  tibble(
    month_num = 1:12,
    month = c("jan", "feb", "mar", "apr", "may", "june", "july", "aug", "sept", "oct", "nov", "dec")
  )

pols_month_df <- pols_month_raw |>
  janitor::clean_names() |> 
  separate(mon, c("year", "month_num", "day"), "-", convert = TRUE) |> 
  mutate(prez = recode(
      prez_dem, 
      "0" = "gop", 
      "1" = "dem")) |> 
  left_join(x = _, y = month_df, by = "month_num") |> 
  select(year, month, everything(), -month_num, -day, -prez_gop, -prez_dem)

```

The pols-month data set describes the number of Democratic and Republican politicians in various political offices at any given time. The data set contains `r nrow(pols_month_df)` observations and `r ncol(pols_month_df)` variables. Variables include `r colnames(pols_month_df)`. Data was collected from `r min(pull(pols_month_df, year))` to `r max(pull(pols_month_df, year))`. 

## Snp Cleaning

Due to the way that the lubridate package converts two digit years into four digit years, we corrected the conversion to ensure twentieth and twenty-first century years were properly captured. Since the other data sets in fivethirtyeight only went up to 2015, we assumed that this was the cutoff date. 

```{r, snp cleaning}
snp_all_dates <- snp_raw |>
  janitor::clean_names() |> 
  mutate(date = lubridate::mdy(date)) |> 
  separate(date, c("year", "month_num", "day"), "-", convert = TRUE) |> 
  left_join(x = _, y = month_df, by = "month_num") |> 
  select(year, month, close, -month_num, -day)

snp_dates_before69 <- snp_all_dates |> 
  filter(!year %in% c(1969:2015)) |> 
  mutate(year = year-100)

snp_dates_after69 <- snp_all_dates |> 
  filter(year %in% c(1969:2015))

snp_combined_df <- bind_rows(snp_dates_before69, snp_dates_after69)

```

The snp data set describes the closing values of the Standard and Poor's stock market index at any given time. The data set contains `r nrow(snp_combined_df)` observations and `r ncol(snp_combined_df)` variables. Variables include `r colnames(snp_combined_df)`. Data was collected from `r min(pull(snp_combined_df, year))` to `r max(pull(snp_combined_df, year))`. 

## Unemployment cleaning

```{r, unemployment cleaning}
unemployment_df <- unemployment_raw |> 
  janitor::clean_names() |> 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "percent_unemployed"
  )

```

The unemployment data set describes the monthly percentage of unemployed individuals in the U.S. from `r min(pull(unemployment_df, year))` to `r max(pull(unemployment_df, year))`. The data set contains `r nrow(unemployment_df)` observations and `r ncol(unemployment_df)` variables. Variables include `r colnames(unemployment_df)`. 

## Joining of FiveThirtyEight datasets 

```{r, merge 538 datasets}
fivethirtyeight_df <- 
  left_join(pols_month_df, snp_combined_df, by = c("year", "month")) |> 
  left_join(x = _, y = unemployment_df, by = c("year", "month"))
  
```

The combined fiverthirtyeight data set compiles contains `r nrow(fivethirtyeight_df)` observations and `r ncol(fivethirtyeight_df)` variables. Variables include `r colnames(fivethirtyeight_df)`. Data spans the years  `r min(pull(fivethirtyeight_df, year))` to `r max(pull(fivethirtyeight_df, year))`.


# Problem 2

## Mr. Trash Wheel Cleaning

Note: The year variable needed to be converted from a character variable to a numeric variable. 

```{r, mr trash cleaning}
mr_trash_df <- mr_trash_raw |> 
  janitor::clean_names() |> 
  rename(volume_cubic_yds = volume_cubic_yards, homes_powered_old = homes_powered) |> 
  mutate(wheel = "mr_trash",
         homes_powered = weight_tons*500/30,
         year = as.numeric(year)) |> 
  select(wheel, everything(), -homes_powered_old)

```

The Mr. Trash Wheel data set contains `r nrow(mr_trash_df)` observations and `r ncol(mr_trash_df)` variables. Variables include `r colnames(mr_trash_df)`. Data was collected from `r min(pull(mr_trash_df, year))` to `r max(pull(mr_trash_df, year))`. During this time, Mr. Trash Wheel collected about `r round(sum(pull(mr_trash_df, weight_tons)))` tons of trash, which powered approximately `r format(round(sum(pull(mr_trash_df, homes_powered))), scientific = FALSE)` homes!

## Prof. Trash Wheel Cleaning

```{r, prof trash cleaning}
prof_trash_df <- prof_trash_raw |> 
  janitor::clean_names() |> 
  rename(volume_cubic_yds = volume_cubic_yards, homes_powered_old = homes_powered) |> 
  mutate(wheel = "prof_trash",
         homes_powered = weight_tons*500/30) |> 
  select(wheel, everything(), -homes_powered_old)

```

The Prof. Trash Wheel data set contains `r nrow(prof_trash_df)` observations and `r ncol(prof_trash_df)` variables. Variables include `r colnames(prof_trash_df)`. Data was collected from `r min(pull(prof_trash_df, year))` to `r max(pull(prof_trash_df, year))`. During this time, Prof. Trash Wheel collected about `r round(sum(pull(prof_trash_df, weight_tons)))` tons of trash, which powered approximately `r round(sum(pull(prof_trash_df, homes_powered)))` homes!

## Gywnnda Trash Wheel Cleaning

```{r, gywnnda trash cleaning}
gwyn_trash_df <- gwyn_trash_raw |> 
  janitor::clean_names() |> 
  rename(volume_cubic_yds = volume_cubic_yards, homes_powered_old = homes_powered) |> 
  mutate(wheel = "gwynnda_trash",
         homes_powered = weight_tons*500/30) |> 
  select(wheel, everything(), -homes_powered_old)

```

The Gwynnda Trash Wheel data set contains `r nrow(gwyn_trash_df)` observations and `r ncol(gwyn_trash_df)` variables. Variables include `r colnames(gwyn_trash_df)`. Data was collected from `r min(pull(gwyn_trash_df, year))` to `r max(pull(gwyn_trash_df, year))`. During this time, Gwynnda Trash Wheel collected about `r round(sum(pull(gwyn_trash_df, weight_tons)))` tons of trash, which powered approximately `r round(sum(pull(gwyn_trash_df, homes_powered)))` homes!


## Joining of Trash Wheel datasets 

```{r, combine trash datasets}
trash_wheels_df <- bind_rows(mr_trash_df, prof_trash_df, gwyn_trash_df)

```

The combined data set contains `r nrow(trash_wheels_df)` observations and `r ncol(trash_wheels_df)` variables. Variables include `r colnames(trash_wheels_df)`. From  `r min(pull(trash_wheels_df, year))` to `r max(pull(trash_wheels_df, year))`, the three trash wheels collected a total of `r round(sum(pull(trash_wheels_df, weight_tons)))` tons of trash, which powered approximately `r format(round(sum(pull(trash_wheels_df, homes_powered))), scientific = FALSE)` homes!

```{r, results = "hide", echo = FALSE}
prof_weight_df <- trash_wheels_df |> 
  filter(wheel == "prof_trash") |> 
  summarize(total_weight = sum(weight_tons))

gywn_cig_df <- trash_wheels_df |> 
  filter(wheel == "gwynnda_trash" & month == "July" & year == "2021") |> 
  summarize(total_cigs = sum(cigarette_butts))
```

Professor Trash Wheel collected a total of `r round(pull(prof_weight_df, total_weight))` tons of trash.

In July 2021, Gwynnda collected a total of `r format(round(pull(gywn_cig_df, total_cigs)), scientific=FALSE)` cigarette butts.

# Problem 3
## MCI_baseline Cleaning

The first two rows of the raw mci_baseline data set contain a header and the variable names. When importing the data set (see above), we skipped the header in the first row and set col_names to TRUE, so we only imported the variable names and observations. **There were no NA values in the data set. The period (".") in the age_at_onset indicated that the participant never developed MCI during the study.**

We recoded the categorical variables of sex and apoe4 status, so the values appeared as the category names instead of numbers.

```{r, mci_baseline_cleaning}
mci_baseline_df <- mci_baseline_raw |> 
  janitor::clean_names() |> 
  mutate(sex = recode(sex, "0" = "female", "1" = "male"),
         apoe4 = recode(apoe4, "0" = "non_carrier", "1" = "carrier")) |> 
  filter(!(current_age == age_at_onset))

removed_participant <- mci_baseline_raw |> 
  janitor::clean_names() |> 
  filter(current_age == age_at_onset)

```


```{r, mci_baseline_analysis, echo = FALSE, results = "hide"}
num_recruited_original <- nrow(mci_baseline_raw)

num_recruited_clean <- nrow(mci_baseline_df)

num_mci <- mci_baseline_df |> 
  filter(!(age_at_onset == ".")) |> 
  nrow()

num_apoe_female <- mci_baseline_df |> 
  filter(sex == "female" & apoe4 == "carrier") |> 
  nrow()

num_female <- mci_baseline_df |> 
  filter(sex == "female") |> 
  nrow()

```

Based on the mci_baseline data set, `r num_recruited_original` participants were recruited into the study. When cleaning the data, we removed participants who did not meet the inclusion criteria for the study (participants had to be free of MCI at baseline). `r num_recruited_original - num_recruited_clean` participant was found to have MCI at baseline, which was was participant `r pull(removed_participant, id)`. `r num_recruited_clean` participants were included in our final analyses. Thus, the cleaned data set contained `r nrow(mci_baseline_df)` observations and `r ncol(mci_baseline_df)` variables

The average baseline age of participants was `r round(mean(pull(mci_baseline_df, current_age)))` years old with a standard deviation of `r round(sd(pull(mci_baseline_df, current_age)))` years.

It was found that `r num_mci` participants (`r round(num_mci / num_recruited_clean *100)`%) developed MCI over the course of the study. Of the `r num_female` women in the study, `r round(num_apoe_female / num_female * 100)`% women were APOE4 carriers.

## MCI_amyloid Cleaning

The first two rows of the raw mci_amyloid data set contain a header and the variable names. When importing the data set (see above), we skipped the header in the first row and set col_names to TRUE, so we only imported the variable names and observations. There are NA values present in the data set. NA values were not removed at this time, because it is important to determine the questions that we would like to answer with the data before identifying if and when NA values should be removed.

```{r, mci_amyloid_cleaning, warning = FALSE}

mci_amyloid_df <- mci_amyloid_raw |> 
  janitor::clean_names() |> 
  rename(id = study_id, time_to_visit_ir = baseline) |> 
  mutate(time_to_visit_ir = as.numeric(time_to_visit_ir), 
         time_2 = as.numeric(time_2),
         time_4 = as.numeric(time_4),
         time_6 = as.numeric(time_6),
         time_8 = as.numeric(time_8)) |> 
  filter(id != pull(removed_participant, id))

```

To clean the mci_amyloid data set, the variables were renamed and the classes of variables were corrected. Here, the time_to_visit_ir:time_8 variables were changed from character to numeric variables. We also removed the one participant who did not meet the study's inclusion criteria based on the mci_baseline data set cleaning. The cleaned mci_amyloid data set contains `r nrow(mci_amyloid_df)` observations and `r ncol(mci_amyloid_df)` variables.

## Joining of MCI data sets

```{r, combine_amyloid_datasets}
mci_no_overlap_df <- anti_join(mci_baseline_df, mci_amyloid_df, join_by(id))

mci_intersection_df <- inner_join(mci_baseline_df, mci_amyloid_df, join_by(id))

write_csv(mci_intersection_df,"data/mci_intersection.csv")

```

It was found that `r nrow(mci_no_overlap_df)` participants were not included in both data sets.


It was found that `r nrow(mci_no_overlap_df)` participants were not included in both data sets.

The combined data set contains `r nrow(mci_intersection_df)` observations and `r ncol(mci_intersection_df)` variables. Variables include `r colnames(mci_intersection_df)`. `r mci_intersection_df |> filter(apoe4 == "carrier") |> nrow()` of the participants in the combined data set were APOE4 carriers.



