---
title: "p8105_hw1_sdc2157"
author: "Stephanie Calluori"
date: 2023-10-04
output: github_document
---

```{r, load_packages, message = FALSE}
library(tidyverse)
library(readxl)
```

# Importing Data
```{r, import_data, results = "hide", message = FALSE}
pols_month_raw <- 
  read_csv("data/pols-month.csv")
head(pols_month_raw)
sum(is.na(pols_month_raw))

snp_raw <-
  read_csv("data/snp.csv")
head(snp_raw)
sum(is.na(snp_raw))

unemployment_raw <-
  read_csv("data/unemployment.csv")
head(unemployment_raw)
sum(is.na(unemployment_raw))

mr_trash_raw <-
  read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = 1, range = "A2:N549")
head(mr_trash_raw)

prof_trash_raw <-
  read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = 2, range = "A2:M96")
head(prof_trash_raw)

gwyn_trash_raw <-
  read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = 4, range = "A2:K108")
head(gwyn_trash_raw)

mci_amyloid_raw <-
  read_csv("data/mci_amyloid.csv", skip = 1, col_names = TRUE)
head(mci_amyloid_raw)
sum(is.na(mci_amyloid_raw))

mci_baseline_raw <- 
  read_csv("data/MCI_baseline.csv", skip = 1, col_names = TRUE)
head(mci_baseline_raw)
sum(is.na(mci_baseline_raw))

```

# Problem 1 - Under REVISION

```{r, pols data exploration, include = FALSE}
pols_month_raw |>
  count(prez_gop)

pols_month_raw |>
  count(prez_dem)
```

```{r, pols cleaning}
month_df = 
  tibble(
    month_num = 1:12,
    month = c("jan", "feb", "mar", "apr", "may", "june", "july", "aug", "sept", "oct", "nov", "dec")
  )

pols_month_df <- pols_month_raw |>
  janitor::clean_names() |> 
  separate(mon, c("year", "month_num", "day"), "-", convert = TRUE) |> 
  mutate(prez = recode(
      prez_dem, 
      "0" = "gop", 
      "1" = "dem")) |> 
  left_join(x = _, y = month_df, by = "month_num") |> 
  select(year, month, everything(), -month_num, -day, -prez_gop, -prez_dem)

head(pols_month_df)

```


```{r, snp cleaning}
snp_df <- snp_raw |>
  janitor::clean_names() |> 
  mutate(date = lubridate::mdy(date)) |> 
  separate(date, c("year", "month_num", "day"), "-", convert = TRUE) |> 
  left_join(x = _, y = month_df, by = "month_num") |> 
  select(year, month, close, -month_num, -day)

head(snp_df)

```


```{r, unemployment cleaning}
unemployment_df <- unemployment_raw |> 
  janitor::clean_names() |> 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "percent_unemployed"
  )

head(unemployment_df)
```


```{r, merge 538 datasets, include = FALSE}
fivethirtyeight_df <- 
  left_join(pols_month_df, snp_df, by = c("year", "month")) |> 
  left_join(x = _, y = unemployment_df, by = c("year", "month"))
  
head(fivethirtyeight_df)
```


# Problem 2

First, we need to clean each of the Trash Wheel datasets.

```{r, mr trash cleaning}
mr_trash_df <- mr_trash_raw |> 
  janitor::clean_names() |> 
  rename(volume_cubic_yds = volume_cubic_yards, homes_powered_old = homes_powered) |> 
  mutate(wheel = "mr_trash",
         homes_powered = weight_tons*500/30,
         year = as.numeric(year)) |> 
  select(wheel, everything(), -homes_powered_old)

str(mr_trash_df)
head(mr_trash_df)
```

The Mr. Trash Wheel data set contains `r nrow(mr_trash_df)` observations and `r ncol(mr_trash_df)` variables. Variables include `r colnames(mr_trash_df)`. Data was collected from `r min(pull(mr_trash_df, year))` to `r max(pull(mr_trash_df, year))`. During this time, Mr. Trash Wheel collected about `r round(sum(pull(mr_trash_df, weight_tons)))` tons of trash, which powered approximately `r format(round(sum(pull(mr_trash_df, homes_powered))), scientific = FALSE)` homes!

```{r, prof trash cleaning}
prof_trash_df <- prof_trash_raw |> 
  janitor::clean_names() |> 
  rename(volume_cubic_yds = volume_cubic_yards, homes_powered_old = homes_powered) |> 
  mutate(wheel = "prof_trash",
         homes_powered = weight_tons*500/30) |> 
  select(wheel, everything(), -homes_powered_old)

str(prof_trash_df)
```

The Prof. Trash Wheel data set contains `r nrow(prof_trash_df)` observations and `r ncol(prof_trash_df)` variables. Variables include `r colnames(prof_trash_df)`. Data was collected from `r min(pull(prof_trash_df, year))` to `r max(pull(prof_trash_df, year))`. During this time, Prof. Trash Wheel collected about `r round(sum(pull(prof_trash_df, weight_tons)))` tons of trash, which powered approximately `r round(sum(pull(prof_trash_df, homes_powered)))` homes!

```{r, gywnnda trash cleaning}
gwyn_trash_df <- gwyn_trash_raw |> 
  janitor::clean_names() |> 
  rename(volume_cubic_yds = volume_cubic_yards, homes_powered_old = homes_powered) |> 
  mutate(wheel = "gwynnda_trash",
         homes_powered = weight_tons*500/30) |> 
  select(wheel, everything(), -homes_powered_old)

str(gwyn_trash_df)
```

The Gwynnda Trash Wheel data set contains `r nrow(gwyn_trash_df)` observations and `r ncol(gwyn_trash_df)` variables. Variables include `r colnames(gwyn_trash_df)`. Data was collected from `r min(pull(gwyn_trash_df, year))` to `r max(pull(gwyn_trash_df, year))`. During this time, Gwynnda Trash Wheel collected about `r round(sum(pull(gwyn_trash_df, weight_tons)))` tons of trash, which powered approximately `r round(sum(pull(gwyn_trash_df, homes_powered)))` homes!


Next, we will combine the 3 Trash Wheel datasets into a single dataset.
```{r, combine trash datasets}
trash_wheels_df <- bind_rows(mr_trash_df, prof_trash_df, gwyn_trash_df)

view(trash_wheels_df)
str(trash_wheels_df)
```

The combined data set contains `r nrow(trash_wheels_df)` observations and `r ncol(trash_wheels_df)` variables. Variables include `r colnames(trash_wheels_df)`. From  `r min(pull(trash_wheels_df, year))` to `r max(pull(trash_wheels_df, year))`, the three trash wheels collected a total of `r round(sum(pull(trash_wheels_df, weight_tons)))` tons of trash, which powered approximately `r format(round(sum(pull(trash_wheels_df, homes_powered))), scientific = FALSE)` homes!

```{r, results = "hide", echo = FALSE}
prof_weight_df <- trash_wheels_df |> 
  filter(wheel == "prof_trash") |> 
  summarize(total_weight = sum(weight_tons))

gywn_cig_df <- trash_wheels_df |> 
  filter(wheel == "gwynnda_trash" & month == "July" & year == "2021") |> 
  summarize(total_cigs = sum(cigarette_butts))
```

Professor Trash Wheel collected a total of `r pull(prof_weight_df, total_weight)` tons of trash.

In July 2021, Gwynnda collected a total of `r pull(gywn_cig_df, total_cigs)` cigarette butts.

# Problem 3
## MCI_Baseline data set
The first two rows of the raw mci_baseline data set contain a header and the variable names. When importing the data set (see above), we skipped the header in the first row and set col_names to TRUE, so we only imported the variable names and observations. There were no NA values in the data set. The period (".") in the age_at_onset indicated that the participant never developed MCI.

We recoded the categorical variables of sex and apoe4 status, so the values appeared as the category names instead of numbers.

```{r}
num_recruited_original <- nrow(mci_baseline_raw)

mci_baseline_df <- mci_baseline_raw |> 
  janitor::clean_names() |> 
  mutate(sex = recode(sex, "0" = "female", "1" = "male"),
         apoe4 = recode(apoe4, "0" = "non_carrier", "1" = "carrier")) |> 
  filter(!(current_age == age_at_onset))

num_recruited_clean <- nrow(mci_baseline_df)

num_mci <- mci_baseline_df |> 
  filter(!(age_at_onset == ".")) |> 
  nrow()

num_apoe_female <- mci_baseline_df |> 
  filter(sex == "female" & apoe4 == "carrier") |> 
  nrow()

num_female <- mci_baseline_df |> 
  filter(sex == "female") |> 
  nrow()
```

`r num_recruited_original` participants were recruited into the study. When cleaning the data, we removed participants who did not meet the inclusion criteria for the study (participants had to be free of MCI at baseline). `r num_recruited_original - num_recruited_clean` participant was found to have MCI at baseline. Thus, `r num_recruited_clean` participants were included in our final analyses. 

The average baseline age of participants was `r round(mean(pull(mci_baseline_df, current_age)))` years old with a standard deviation of `r round(sd(pull(mci_baseline_df, current_age)))` years.

It was found that `r num_mci` participants (`r round(num_mci / num_recruited_clean *100)`%) developed MCI over the course of the study. Of the `r num_female` women in the study, `r round(num_apoe_female / num_female * 100)`% women were APOE4 carriers.

#MCI_Amyloid data set

```{r}

str(mci_amyloid_raw)
nrow(mci_amyloid_raw)

mci_amyloid_df <- mci_amyloid_raw |> 
  janitor::clean_names() |> 
  rename(id = study_id, time_to_visit_ir = baseline) |> 
  mutate(time_to_visit_ir = as.numeric(time_to_visit_ir), 
         time_2 = as.numeric(time_2),
         time_4 = as.numeric(time_4),
         time_6 = as.numeric(time_6),
         time_8 = as.numeric(time_8))

str(mci_amyloid_df)
nrow(mci_amyloid_df)
  
```

```{r}
mci_no_overlap_df <- anti_join(mci_baseline_df, mci_amyloid_df, join_by(id))

mci_intersection_df <- inner_join(mci_baseline_df, mci_amyloid_df, join_by(id))

write_csv(mci_intersection_df,"data/mci_intersection.csv")

```





